Accounting for uncertainty in simulation greenhouse gas emissions from dairy cattle in Kenya

### 2022-06-01 Manuscript_analyses.Rmd
some polishing

---
title: "Accounting for uncertainty in simulation greenhouse gas emissions from dairy cattle in Kenya"
author: "Eike Luedeling, Cory Whitney, Todd Rosenstock"
bibliography:
 - references/packages.bib
 - references/papers.bib
output: html_document
---

@@

# Load emission calculation function

The formulas come from Chapter 10 of the *2006 IPCC Guidelines for National Greenhouse Gas Inventories* [@dong_ipcc_2006] (Volume 4 - Agriculture, Forestry and Other Land Use). The document can be found on [the IPCC website](https://www.ipcc-nggip.iges.or.jp/public/2006gl/pdf/4_Volume4/V4_10_Ch10_Livestock.pdf). The chapter is part of the book *Good Practice Guidance and Uncertainty Management in National Greenhouse Gas Inventories* [@ipcc_good_2021], also on the [the IPCC website](https://www.ipcc-nggip.iges.or.jp/public/gp/english/). Many variables are based on the work done in *Variation in the carbon footprint of milk production on smallholder dairy farms in central Kenya* [@wilkes_variation_2020] and in *Methods and guidance to support MRV of livestock emissions* [@wilkes_methods_2019]. We run the model with the decisionSupport package [@R-decisionSupport].

@@

This equation takes a wide range of inputs. Many of these are usually not known precisely, yet the classic IPCC equation does not account for potential errors and uncertainties. Here, we express this uncertainty by defining distributions for each variable. The following table shows all the variables, as well as examples of distributions that may describe their quantities. Distributions are defined according to the distribution shape, as well as upper and lower bounds of 90% confidence intervals.

```{r load_data}

### # This project will be stored in relative file paths (i.e. same or lower folders)

@@

options(knitr.kable.NA = '', encoding = "Windows-1252")
 read.csv(input_table)[,c(1:2,4,7,5:6)]   %>%
    kableExtra::kbl(digits = 2)  %>%
    kableExtra::kable_classic_2(full_width = F)

@@

# Monte Carlo simulation

@@

# Running a Monte Carlo simulation with scenarios

@@

The `scenario_mc` function in the `decisionSupport` package [@R-decisionSupport] consists of a wrapper around the `mcSimulation` function. The function is able to modify the input table according to a scenario file we provide. This file contains information on which variables should be modified for each scenario and how many model runs should be included in the simulation. The function is essentially the same as the `mcSimulation` function, but with an additional `scenarios` option which imports the contents of a .csv file.

The .csv file for the `scenarios` option specifies variables that should be modified for each of the scenarios, as well as the number of model runs for it. Each column beyond the first two columns in the file contain information for a particular scenario. This is restricted to the parameters that *should be changed* for this scenario.

@@

# Household-based scenarios

@@

@@

```{r}
farm5_emissions$y[,"CO2eq_per_milk"]<-farm5_emissions$y$pc_on_farm/farm5_emissions$y$pc_milk_yield

@@

# Population-wide emission density of milk production

@@

```{r, warning=FALSE, message=FALSE}
#uncertainty::varscatter(in_var = GHG_simulation_scenarios$y$pc_milk_yield,
#                        out_var = GHG_simulation_scenarios$y$pc_on_farm,
#                        xlab = expression(Total~milk~yield~(kg~head^{-1}~year^{-1})),
#                        ylab = expression(Total~emissions~(kg~CO[2]-eq~head^{-1}~year^{-1})))

plotdata<-data.frame(in_var=GHG_simulation_scenarios$y$pc_milk_yield,
                     out_var=GHG_simulation_scenarios$y$pc_on_farm,
                     Group="just one")

@@

# Emissions intensity of milk production

@@

#uncertainty::varscatter(in_var = dairy_households$y$pc_milk_yield,
#                        out_var = dairy_households$y$CO2eq_per_milk,
#                        xlab = expression(Total~milk~yield~(kg~head^{-1}~year^{-1})),
#                        ylab = expression(Total~emissions~intensity~of~milk~production~(kg~CO[2]-eq~kg^{-1})))

@@

# Probability of GHG emissions for a given milk yield

@@

#graphics::plot(x = slice$slice$Output_values, 
#                 y = slice$slice$Relative_probability, 
#                 type = "l", 
#                 ylab = "Relative probability", 
#                 xlab = "Emissions intensity", 
#                 col = "seagreen", 
#                 lwd = 2)

@@

# Probability of an increase in milk yield being associated with an increase in GHG emission intensity

@@

# Testing interventions - strategy

@@

# Testing interventions

@@

## Breed intervention

@@

# Compare outcomes of breed intervention with baseline

@@

ggplot(data=impact_of_breed,aes(x=milk_change,y=emissions_intensity_change)) + geom_point() + theme_bw() + ylab("Change in emissions intensity (kg CO2-eq kg-1)") + xlab("Change in per capita milk yield (kg)")

ggsave("Fig_10_breed_reductions.png", width=7,height=5)

ggplot(data = decreased_emissions_probability, aes(Base_yield, yield_increase)) + geom_contour_filled(aes(z=probability)) +
  stat_contour(breaks = c(0.9),aes(z=probability),col="black") +
  geom_point(data=impact_of_breed[which(impact_of_breed$milk_change>0),],aes(x=baseline,y=milk_change)) +
  ggtitle("Probability that emission intensity reductions occurred") + 
  xlab("Baseline milk yield per head and year (kg)") +
  ylab("Milk yield increase per head and year (kg)") 
ggsave("Fig_11_breed_probability_of_intensity_reduction.png", width=7,height=5)

@@

## Herd structure

### Retiring unproductive bulls

@@

#### Evaluate impact of "Retiring unproductive bulls" intervention

@@

```{r, fig.show="hold", fig.cap="No bulls intervention (left) vs. baseline (right)", message=FALSE, warning=FALSE}

@@

#uncertainty::varscatter(in_var = nobull_res$y$pc_milk_yield,
#                        out_var = nobull_res$y$CO2eq_per_milk,
#                        xlab = expression(Total~milk~yield~(kg~head^{-1}~year^{-1})),
#                        ylab = expression(Total~emissions~(kg~CO[2]-eq~head^{-1}~year^{-1})))

#uncertainty::varscatter(in_var = GHG_simulation_scenarios$y$pc_milk_yield,
#                        out_var = GHG_simulation_scenarios$y$CO2eq_per_milk,
#                        xlab = expression(Total~milk~yield~(kg~head^{-1}~year^{-1})),
#                        ylab = expression(Total~emissions~(kg~CO[2]-eq~head^{-1}~year^{-1})))

#uncertainty::varkernel(in_var = nobull_res$y$pc_milk_yield,
#                        out_var = nobull_res$y$pc_on_farm,
#                        xlab = expression(Total~milk~yield~(kg~head^{-1}~year^{-1})),
#                        ylab = expression(Total~emissions~(kg~CO[2]-eq~head^{-1}~year^{-1})))

#uncertainty::varkernel(in_var = GHG_simulation_scenarios$y$pc_milk_yield,
#                        out_var = GHG_simulation_scenarios$y$pc_on_farm,
#                        xlab = expression(Total~milk~yield~(kg~head^{-1}~year^{-1})),
#                        ylab = expression(Total~emissions~(kg~CO[2]-eq~head^{-1}~year^{-1})))

@@

dens_nobull<-ggplot(plotdata_nobull, aes(x = in_var, y = out_var)) +
  stat_density_2d(
    geom = "raster",
    aes(fill = after_stat(density)),
    contour = FALSE
    ) +
  scale_fill_viridis_c() + 
  labs(y=expression(Total~emissions~intensity~of~milk~production~(kg~CO[2]*'-eq'~kg^{-1})),
       x=expression(Total~milk~yield~(kg~head^{-1}~year^{-1}))) +
  theme_bw()

dens_baseline<-ggplot(plotdata_baseline, aes(x = in_var, y = out_var)) +
  stat_density_2d(
    geom = "raster",
    aes(fill = after_stat(density)),
    contour = FALSE
    ) +
    scale_y_continuous(position = "right") +
  scale_fill_viridis_c() + 
  labs(y=expression(Total~emissions~intensity~of~milk~production~(kg~CO[2]*'-eq'~kg^{-1})),
       x=expression(Total~milk~yield~(kg~head^{-1}~year^{-1}))) +
  theme_bw()

plot_nobull + plot_baseline + dens_nobull + dens_baseline +
  plot_layout(ncol = 2, nrow = 2) + guides(fill="none")

ggsave("Fig_12_no_bulls_scatter_per_head_milk_emissions_intensity.png",width=9,height=9)

@@

### Reducing unproductive replacement males

@@

# Compare outcomes of herd intervention with baseline

## No bulls scenario

@@

impact_of_nobull<-data.frame(baseline=dairy_households$y$pc_milk_yield,breed_milk=nobull_res$y$pc_milk_yield,CO2_intensity_baseline= dairy_households$y$CO2eq_per_milk,CO2_intensity_breed=nobull_res$y$CO2eq_per_milk)
impact_of_nobull[,"milk_change"]<-round(impact_of_nobull$breed_milk-impact_of_nobull$baseline,2)
impact_of_nobull[,"emissions_intensity_change"]<-round(impact_of_nobull$CO2_intensity_breed-impact_of_nobull$CO2_intensity_baseline ,3)

ggplot(data=impact_of_nobull,aes(x=milk_change,y=emissions_intensity_change)) + geom_point() + theme_bw() + ylab("Change in emissions intensity (kg CO2-eq kg-1)") + xlab("Change in per capita milk yield (kg)")

ggplot(data = decreased_emissions_probability, aes(Base_yield, yield_increase)) + geom_contour_filled(aes(z=probability)) +
  stat_contour(breaks = c(0.9),aes(z=probability),col="black") +
  geom_point(data=impact_of_nobull[which(impact_of_nobull$milk_change>0),],aes(x=baseline,y=milk_change)) +
  ggtitle("Probability that emission intensity reductions occurred") + 
  xlab("Baseline milk yield per head and year (kg)") +
  ylab("Milk yield increase per head and year (kg)")

ggsave("Fig_13_impact_of_nobull_reductions.png")

@@

## Fewer replacement males scenario

@@

impact_of_lessReplace<-data.frame(baseline=dairy_households$y$pc_milk_yield,breed_milk=lessReplace_res$y$pc_milk_yield,CO2_intensity_baseline= dairy_households$y$CO2eq_per_milk,CO2_intensity_breed=lessReplace_res$y$CO2eq_per_milk)
impact_of_lessReplace[,"milk_change"]<-round(impact_of_lessReplace$breed_milk-impact_of_lessReplace$baseline,2)
impact_of_lessReplace[,"emissions_intensity_change"]<-round(impact_of_lessReplace$CO2_intensity_breed-impact_of_lessReplace$CO2_intensity_baseline ,3)

ggplot(data=impact_of_lessReplace,aes(x=milk_change,y=emissions_intensity_change)) + geom_point() + theme_bw() + ylab("Change in emissions intensity (kg CO2-eq kg-1)") + xlab("Change in per capita milk yield (kg)")

ggplot(data = decreased_emissions_probability, aes(Base_yield, yield_increase)) + geom_contour_filled(aes(z=probability)) +
  stat_contour(breaks = c(0.9),aes(z=probability),col="black") +
  geom_point(data=impact_of_lessReplace[which(impact_of_lessReplace$milk_change>0),],aes(x=baseline,y=milk_change)) +
  ggtitle("Probability that emission intensity reductions occurred") + 
  xlab("Baseline milk yield per head and year (kg)") +
  ylab("Milk yield increase per head and year (kg)")

ggsave("Fig_14_lessReplace_reductions.png")

>>>>

---
title: "Accounting for uncertainty in simulation greenhouse gas emissions from dairy cattle in Kenya"
author: "Eike Luedeling, Cory Whitney, Todd Rosenstock"
bibliography:
 - references/packages.bib
 - references/papers.bib
output:
  html_document:
    theme: united
    toc: true
    toc_float: true
    number_sections: true
---

@@

## Load emission calculation function

Greenhouse gas emissions from livestock are calculated according to Chapter 10 of the *2006 IPCC Guidelines for National Greenhouse Gas Inventories* [@dong_ipcc_2006] (Volume 4 - Agriculture, Forestry and Other Land Use). The document can be found on [the IPCC website](https://www.ipcc-nggip.iges.or.jp/public/2006gl/pdf/4_Volume4/V4_10_Ch10_Livestock.pdf). The chapter is part of the book *Good Practice Guidance and Uncertainty Management in National Greenhouse Gas Inventories* [@ipcc_good_2021], also on the [the IPCC website](https://www.ipcc-nggip.iges.or.jp/public/gp/english/). Many variables are based on the work done in *Variation in the carbon footprint of milk production on smallholder dairy farms in central Kenya* [@wilkes_variation_2020] and in *Methods and guidance to support MRV of livestock emissions* [@wilkes_methods_2019]. We also draw on recommendations provided in Chapter 10 of Volume 4 of the *2019 Refinement to the 2006 IPCC Guidelines for National Greenhouse Gas Inventories* [@ipcc_2019], which is also accessible via [the IPCC website](https://www.ipcc-nggip.iges.or.jp/public/2019rf/vol4.html).

@@

This equation takes a wide range of inputs. Many of these are usually not known precisely, yet the classic IPCC equation does not account for potential errors and uncertainties. Here, we express this uncertainty by defining distributions for each variable. The following table (**data/livestock_ghg_input_table.csv**) shows all the variables, as well as examples of distributions that may describe their quantities. Distributions are defined according to the distribution shape, as well as upper and lower bounds of 90% confidence intervals.

```{r load_data, include=FALSE}

@@

options(knitr.kable.NA = '', encoding = "Windows-1252")
 read.csv(input_table)[,c(1:2,4,7,5:6)]   %>%
    kableExtra::kbl(digits = 2)  %>%
    kableExtra::kable_classic_2(full_width = F) %>% 
 scroll_box(height = "300px")

@@

## Monte Carlo simulation

@@

## Running a Monte Carlo simulation with scenarios

@@

The `scenario_mc` function in the `decisionSupport` package [@R-decisionSupport] consists of a wrapper around the `mcSimulation` function. The function can modify the input table according to a scenario file we provide. This file contains information on which variables should be modified for each scenario and how many model runs should be included in the simulation. The function is essentially the same as the `mcSimulation` function, but with an additional `scenarios` option which imports the contents of a .csv file.

The .csv file for the `scenarios` option specifies variables that should be modified for each of the scenarios, as well as the number of model runs for it. Each column beyond the first two columns in the file contains information for a particular scenario. This is restricted to the parameters that *should be changed* for this scenario.

@@

## Household-based scenarios

@@

## Uncertainty at the farm scale

@@

```{r, warning=FALSE, message=FALSE}
farm5_emissions$y[,"CO2eq_per_milk"]<-farm5_emissions$y$pc_on_farm/farm5_emissions$y$pc_milk_yield

@@

## Population-wide emission density of milk production

@@

```{r, warning=FALSE, message=FALSE}
plotdata<-data.frame(in_var=GHG_simulation_scenarios$y$pc_milk_yield,
                     out_var=GHG_simulation_scenarios$y$pc_on_farm,
                     Group="just one")

@@

## Emissions intensity of milk production

@@

@@

# Milk yield as an indicator of GHG emissions intensity

## Probability of GHG emissions for a given milk yield

@@

@@

## Probability of an increase in milk yield being associated with an increase in GHG emission intensity

@@

# Methodology for testing interventions

@@

@@

# Emission reductions through higher-performance breeds

@@

## Emissions intensity response to introduction of high-performance breeds

@@

breed_scatter<-ggplot(data=impact_of_breed,
                      aes(x=milk_change,y=emissions_intensity_change)) +
  geom_point() +
  theme_bw() +
  ggtitle("Milk yield and emissions response") + 
  labs(y=expression(Change~'in'~emission~intensity~(kg~CO[2]*'-eq'~kg^-1)),
       x="Change in per capita milk yield (kg)")

breed_probs<-ggplot(data = decreased_emissions_probability,
                    aes(Base_yield, yield_increase)) +
  geom_contour_filled(aes(z=probability)) +
  stat_contour(breaks = c(0.9),
               aes(z=probability),col="black") +
  geom_point(data=impact_of_breed[which(impact_of_breed$milk_change>0),],aes(x=baseline,y=milk_change)) +
  ggtitle("Probability that emission intensity reductions occurred") + 
  labs(y=expression(Milk~yield~increase~(kg~head^{-1}~year^{-1})),
       x=expression(Baseline~milk~yield~(kg~head^{-1}~year^{-1})),
       fill="Probability")

breed_scatter + breed_probs +
  plot_layout(ncol = 1, nrow = 2)

ggsave("Fig_10_breed_reductions.png", width=9,height=9)

@@

# Herd structure adjustments

## Retiring unproductive bulls

@@

## Emissions intensity response to retiring of unproductive bulls

@@

```{r, message=FALSE, warning=FALSE}

@@

@@

dens_nobull<-ggplot(plotdata_nobull, aes(x = in_var, y = out_var)) +
  stat_density_2d(
    geom = "raster",
    aes(fill = after_stat(density)),
    contour = FALSE
    ) +
  scale_fill_viridis_c() + 
  labs(y=expression(Emissions~intensity~of~milk~production~(kg~CO[2]*'-eq'~kg^{-1})),
       x=expression(Total~milk~yield~(kg~head^{-1}~year^{-1}))) +
  theme_bw() + guides(fill="none")

dens_baseline<-ggplot(plotdata_baseline, aes(x = in_var, y = out_var)) +
  stat_density_2d(
    geom = "raster",
    aes(fill = after_stat(density)),
    contour = FALSE
    ) +
    scale_y_continuous(position = "right") +
  scale_fill_viridis_c() + 
  labs(y=expression(Emissions~intensity~of~milk~production~(kg~CO[2]*'-eq'~kg^{-1})),
       x=expression(Total~milk~yield~(kg~head^{-1}~year^{-1}))) +
  theme_bw() + guides(fill="none")

plot_nobull + plot_baseline + dens_nobull + dens_baseline +
  plot_layout(ncol = 2, nrow = 2)

ggsave("Fig_11_no_bulls_scatter_per_head_milk_emissions_intensity.png",width=9,height=9)

@@

## Reducing unproductive replacement males

@@

## Herd interventions vs. baseline scenario

### No bulls scenario

@@

impact_of_nobull<-data.frame(baseline=dairy_households$y$pc_milk_yield,
                             breed_milk=nobull_res$y$pc_milk_yield,
                             CO2_intensity_baseline= dairy_households$y$CO2eq_per_milk,
                             CO2_intensity_breed=nobull_res$y$CO2eq_per_milk)
impact_of_nobull[,"milk_change"]<-
  round(impact_of_nobull$breed_milk-impact_of_nobull$baseline,2)
impact_of_nobull[,"emissions_intensity_change"]<-
  round(impact_of_nobull$CO2_intensity_breed-impact_of_nobull$CO2_intensity_baseline ,3)

nobull_scatter<-ggplot(data=impact_of_nobull,
                       aes(x=milk_change,y=emissions_intensity_change)) +
  geom_point() +
  theme_bw() +
  ggtitle("Milk yield and emissions response") + 
  labs(y=expression(Change~'in'~emission~intensity~(kg~CO[2]*'-eq'~kg^-1)),
       x="Change in per capita milk yield (kg)")

nobull_probs<-ggplot(data = decreased_emissions_probability,
                     aes(Base_yield, yield_increase)) +
  geom_contour_filled(aes(z=probability)) +
  stat_contour(breaks = c(0.9),aes(z=probability),col="black") +
  geom_point(data=impact_of_nobull[which(impact_of_nobull$milk_change>0),],
             aes(x=baseline,y=milk_change)) +
  ggtitle("Probability that emission intensity reductions occurred") + 
  labs(y=expression(Milk~yield~increase~(kg~head^{-1}~year^{-1})),
       x=expression(Baseline~milk~yield~(kg~head^{-1}~year^{-1})),
       fill="Probability")

nobull_scatter + nobull_probs +
  plot_layout(ncol = 1, nrow = 2)

ggsave("Fig_12_impact_of_nobull_reductions.png", width=9,height=5)

@@

### Fewer replacement males scenario

@@

impact_of_lessReplace<-data.frame(baseline=dairy_households$y$pc_milk_yield,
                                  breed_milk=lessReplace_res$y$pc_milk_yield,
                                  CO2_intensity_baseline= dairy_households$y$CO2eq_per_milk,
                                  CO2_intensity_breed=lessReplace_res$y$CO2eq_per_milk)
impact_of_lessReplace[,"milk_change"]<-
  round(impact_of_lessReplace$breed_milk-impact_of_lessReplace$baseline,2)
impact_of_lessReplace[,"emissions_intensity_change"]<-
  round(impact_of_lessReplace$CO2_intensity_breed-impact_of_lessReplace$CO2_intensity_baseline ,3)

lessRep_scatter<-ggplot(data=impact_of_lessReplace,
                        aes(x=milk_change,y=emissions_intensity_change)) +
  geom_point() +
  theme_bw() +
  ggtitle("Intervention impact on milk yield and emissions") + 
  labs(y=expression(Change~'in'~emission~intensity~(kg~CO[2]*'-eq'~kg^-1)),
       x="Change in per capita milk yield (kg)")

lessRep_probs<-ggplot(data = decreased_emissions_probability,
                      aes(Base_yield, yield_increase)) +
  geom_contour_filled(aes(z=probability)) +
  stat_contour(breaks = c(0.9),aes(z=probability),col="black") +
  geom_point(data=impact_of_lessReplace[which(impact_of_lessReplace$milk_change>0),],
             aes(x=baseline,y=milk_change)) +
  ggtitle("Probability that emission intensity reductions occurred") + 
  labs(y=expression(Milk~yield~increase~(kg~head^{-1}~year^{-1})),
       x=expression(Baseline~milk~yield~(kg~head^{-1}~year^{-1})),
       fill="Probability")

lessRep_scatter + lessRep_probs +
  plot_layout(ncol = 1, nrow = 2)

ggsave("Fig_13_lessReplace_reductions.png", width=9,height=9)
